---
title: "Lab 2: Differential expression Part 2"
author: "Jean-Philippe Fortin"
date: "April 10, 2014"
output: ioslides_presentation
---

## This week  
The lab on Friday will be an introduction to the cluster. These slides provide solutions to the previous lab exercises. 

## Solutions to Lab 1 Exercise 1
You were asked to 

- Read about loess/lowess normalization. 
- Normalize samples 1 and 2 using the lowess curve that we computed
- How does the MA plot look like after normalization? Does it look better than quantile normalization?

```{r,echo,results='hide', message=FALSE, warning=FALSE}
library(affy)
library(affydata)
data(Dilution)
raw <- pm(Dilution)
log.raw <- log2(raw)
avg  <- log.raw[,1] + log.raw[,2]
diff <-  log.raw[,1] - log.raw[,2]
```

## Solutions to Lab 1 Exercise 1
To fit the lowess curve:
```{r}
lowess.curve <- lowess(x = avg, y = diff, f = 0.05)
names(lowess.curve)
```
**x** corresponds to the sorted variable **avg**, and **y** corresponds to the predicted value of **diff** using the lowess curve. For each value of the vector **avg**, we need to find its corresponding value in the vector **x**. The function **match()** does that for us (see **?match**).


## Solutions to Lab 1 Exercise 1
Lowess normalization:
```{r}
y.predicted <- lowess.curve$y[match(avg, lowess.curve$x)]
new.diff <- diff - y.predicted
plot(avg, new.diff, pch=20, cex=0.5) # MA plot
abline(h=0,col="deepskyblue3", lwd=4)
```

## Solutions to Lab 1 Exercise 1
We have corrected the Avg and Diff values; how can we retrieve the corrected intensities of Array 1 and Array 2? First, we used the bivariate transformation
$$ avg = x_1+x_2$$
$$ diff = x_1-x_2$$
Its inverse transformation is
$$ x_1 = \frac{avg+diff}{2}$$
$$ x_2 = \frac{avg-diff}{2}$$

## Solutions to Lab 1 Exercise 1
Then
```{r}
new.array1 <- (avg+new.diff)/2
new.array2 <- (avg-new.diff)/2
plot(new.array1, new.array2, pch=20, cex=0.5)
abline(a=0,b=1,col="deepskyblue3", lwd=4)
```

## Solutions to Lab 1 Exercise 2
We've done it in class. If you'd like to see it one more time, please email me.

## Solutions to Lab 1 Exercise 3
You were asked to 

- Repeat the differential expression analysis, but this time testing for the difference between samples from scanner 1 and samples from scanner 2 (see phenotype information)
- Produce a p-values (computed with **eBayes**) histogram 
- Repeat the analysis but this time with data that haven't been background corrected and that are not normalized. Use the median polish algorithm for the probeset summarization. Use the function **expresso()** in the **affy** package to produce such an expression matrix. 
- What can you tell about the p-values produced with the RMA-corrected dataset? What can you tell about the p-values produced with the uncorrected data? 

Let's do it.

## Solutions to Lab 1 Exercise 2
First, let's create our group variable; we want to use the scanner variable this time:
```{r}
pd <- pData(Dilution)
groups <- as.factor(pd$scanner)
```
and create the corresponding design matrix:
```{r}
design.matrix <- model.matrix(~groups)
colnames(design.matrix) <- c("Intercept","Scanner")
```
In the first question , we want to use the RMA-corrected expression matrix:
```{r}
e <- exprs(rma(Dilution))
```


## Solutions to Lab 1 Exercise 2
Then we load **limma** and run a linear model:
```{r, warning=FALSE, message=FALSE}
library(limma)
fit <- lmFit(e, design.matrix)
```
and compute the moderated t-statistics using Empirical Bayes:
```{r}
fit2 <- eBayes(fit)
```
What does the histogram of the p-values look like?


## Solutions to Lab 1 Exercise 2
```{r}
hist(fit2$p.value[,2])
```
What can you tell?

## Solutions to Lab 1 Exercise 2
Let's do it one more time, but without using background correcton and quantile normalization (using the function **expresso**):
```{r}
e2 <- expresso(Dilution, bg.correct = FALSE, normalize = FALSE, 
               pmcorrect.method="pmonly", summary.method = "medianpolish")
fit3 <- eBayes(lmFit(e2, design.matrix))
```

## Solutions to Lab 1 Exercise 2
Do you think this histogram looks good? What's going on?
```{r}
hist(fit3$p.value[,2])
```

## Next
At the beginning of next lab we will

- map the probesets to real genes
- create a nice table of differentially expressed genes ready for publication

